---
title: "Heart Attack Risk Assessment"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Heart Attack Risk Assessment |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/fajobgiua/heart-attack-risk-assessment-dataset/data\>*

### Reference:

*\<Job Giua Fa (2023). Heart Attack Risk Assessment Dataset. Kaggle. https://www.kaggle.com/datasets/fajobgiua/heart-attack-risk-assessment-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*


# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r Loading the dataset}
# Load dataset using read.csv
HeartRiskData <- read.csv("Heart_Attack_Risk_Levels_Dataset.csv", colClasses = c(
  Age = "numeric",
  Gender = "factor",
  Heart_rate = "numeric",
  Systolic_blood_pressure = "numeric",
  Diastolic_blood_pressure = "numeric",
  Blood_sugar = "numeric",
  CK_MB = "numeric",
  Troponin = "numeric",
  Result = "factor",
  Risk_Level = "factor"
))

# Preview structure and first few rows
str(HeartRiskData)
head(HeartRiskData)
View(HeartRiskData)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Frequency count for categorical variables (e.g., Gender, Result, Risk_Level)
table(HeartRiskData$Gender)
table(HeartRiskData$Result)
table(HeartRiskData$Risk_Level)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean, Median for numeric variables
mean(HeartRiskData$Age, na.rm = TRUE)
median(HeartRiskData$Age, na.rm = TRUE)

mean(HeartRiskData$Heart_rate, na.rm = TRUE)
median(HeartRiskData$Heart_rate, na.rm = TRUE)

mean(HeartRiskData$Systolic_blood_pressure, na.rm = TRUE)
median(HeartRiskData$Systolic_blood_pressure, na.rm = TRUE)

mean(HeartRiskData$Diastolic_blood_pressure, na.rm = TRUE)
median(HeartRiskData$Diastolic_blood_pressure, na.rm = TRUE)

mean(HeartRiskData$Blood_sugar, na.rm = TRUE)
median(HeartRiskData$Blood_sugar, na.rm = TRUE)

mean(HeartRiskData$CK_MB, na.rm = TRUE)
median(HeartRiskData$CK_MB, na.rm = TRUE)

mean(HeartRiskData$Troponin, na.rm = TRUE)
median(HeartRiskData$Troponin, na.rm = TRUE)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Standard Deviation and Range for numeric variables
sd(HeartRiskData$Age, na.rm = TRUE)
sd(HeartRiskData$Heart_rate, na.rm = TRUE)
sd(HeartRiskData$Systolic_blood_pressure, na.rm = TRUE)
sd(HeartRiskData$Diastolic_blood_pressure, na.rm = TRUE)
sd(HeartRiskData$Blood_sugar, na.rm = TRUE)
sd(HeartRiskData$CK_MB, na.rm = TRUE)
sd(HeartRiskData$Troponin, na.rm = TRUE)

# Range of variables
range(HeartRiskData$Age, na.rm = TRUE)
range(HeartRiskData$Heart_rate, na.rm = TRUE)
range(HeartRiskData$Systolic_blood_pressure, na.rm = TRUE)
range(HeartRiskData$Diastolic_blood_pressure, na.rm = TRUE)
range(HeartRiskData$Blood_sugar, na.rm = TRUE)
range(HeartRiskData$CK_MB, na.rm = TRUE)
range(HeartRiskData$Troponin, na.rm = TRUE)

```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation between numeric variables
cor(HeartRiskData$Age, HeartRiskData$Heart_rate, use = "complete.obs")
cor(HeartRiskData$Systolic_blood_pressure, HeartRiskData$Diastolic_blood_pressure, use = "complete.obs")
cor(HeartRiskData$Blood_sugar, HeartRiskData$CK_MB, use = "complete.obs")
cor(HeartRiskData$Troponin, HeartRiskData$Heart_rate, use = "complete.obs")

# Cross-tabulation for categorical variables
table(HeartRiskData$Result, HeartRiskData$Risk_Level)
```

## ANOVA
```{r ANOVA}
# ANOVA: Test if Age differs significantly across Risk_Level groups
anova_age_risk <- aov(Age ~ Risk_Level, data = HeartRiskData)
summary(anova_age_risk)

# ANOVA: Test if Heart_rate differs significantly across Risk_Level groups
anova_heart_rate_risk <- aov(Heart_rate ~ Risk_Level, data = HeartRiskData)
summary(anova_heart_rate_risk)

# ANOVA: Test if Systolic_blood_pressure differs significantly across Risk_Level groups
anova_systolic_risk <- aov(Systolic_blood_pressure ~ Risk_Level, data = HeartRiskData)
summary(anova_systolic_risk)

# ANOVA: Test if Blood_sugar differs significantly across Risk_Level groups
anova_blood_sugar_risk <- aov(Blood_sugar ~ Risk_Level, data = HeartRiskData)
summary(anova_blood_sugar_risk)

# ANOVA: Test if CK_MB differs significantly across Risk_Level groups
anova_ck_mb_risk <- aov(CK_MB ~ Risk_Level, data = HeartRiskData)
summary(anova_ck_mb_risk)

# ANOVA: Test if Troponin differs significantly across Risk_Level groups
anova_troponin_risk <- aov(Troponin ~ Risk_Level, data = HeartRiskData)
summary(anova_troponin_risk)

# Normality check
plot(anova_age_risk, 2)  # Q-Q plot for Age vs Risk_Level

# Homogeneity of variance
bartlett.test(Age ~ Risk_Level, data = HeartRiskData)
```

## Plots
```{r Plots}
# Histogram for Age
hist(HeartRiskData$Age, 
     main = "Distribution of Age", 
     xlab = "Age", 
     col = "lightblue", 
     border = "black")

# Boxplot for Heart_rate
boxplot(HeartRiskData$Heart_rate, 
        main = "Heart Rate Distribution", 
        ylab = "Heart Rate", 
        col = "lightgreen")

# Histogram for Systolic Blood Pressure
hist(HeartRiskData$Systolic_blood_pressure, 
     main = "Distribution of Systolic Blood Pressure", 
     xlab = "Systolic Blood Pressure", 
     col = "lightcoral", 
     border = "black")

# Bar plot for Risk_Level
barplot(table(HeartRiskData$Risk_Level), 
        main = "Distribution of Risk Level", 
        xlab = "Risk Level", 
        ylab = "Frequency", 
        col = c("lightblue", "lightgreen", "lightcoral"))

# Density plot for Blood_sugar
plot(density(HeartRiskData$Blood_sugar), 
     main = "Density Plot of Blood Sugar", 
     xlab = "Blood Sugar", 
     col = "purple")

# Pairs plot for numeric variables
pairs(HeartRiskData[, c("Age", "Heart_rate", "Systolic_blood_pressure", 
                        "Diastolic_blood_pressure", "Blood_sugar", "CK_MB", "Troponin")], 
      main = "Pairs Plot for Numeric Variables", 
      pch = 19, col = HeartRiskData$Risk_Level)

# Load necessary library for heatmap
library(corrplot)

# Calculate correlation matrix
cor_matrix <- cor(HeartRiskData[, c("Age", "Heart_rate", "Systolic_blood_pressure", 
                                    "Diastolic_blood_pressure", "Blood_sugar", "CK_MB", "Troponin")])

# Plot the correlation matrix
corrplot(cor_matrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45, 
         main = "Correlation Matrix of Numeric Variables")

# Scatter plot for Age vs Heart_rate, colored by Risk_Level
plot(HeartRiskData$Age, HeartRiskData$Heart_rate, 
     main = "Scatter Plot of Age vs Heart Rate", 
     xlab = "Age", ylab = "Heart Rate", 
     col = HeartRiskData$Risk_Level, pch = 19)
legend("topright", legend = levels(HeartRiskData$Risk_Level), 
       col = 1:length(levels(HeartRiskData$Risk_Level)), 
       pch = 19)

# Boxplot of Heart_rate by Risk_Level
boxplot(Heart_rate ~ Risk_Level, data = HeartRiskData, 
        main = "Heart Rate by Risk Level", 
        xlab = "Risk Level", ylab = "Heart Rate", 
        col = c("lightblue", "lightgreen", "lightcoral"))

# Violin plot for Systolic_blood_pressure by Risk_Level
library(ggplot2)
ggplot(HeartRiskData, aes(x = Risk_Level, y = Systolic_blood_pressure, fill = Risk_Level)) +
  geom_violin(trim = FALSE) +
  labs(title = "Violin Plot of Systolic Blood Pressure by Risk Level",
       x = "Risk Level", y = "Systolic Blood Pressure") +
  theme_minimal()


```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check if there are any missing values in the entire dataset
missing_values <- sum(is.na(HeartRiskData))
cat("Total missing values in the dataset:", missing_values, "\n")

# Check for missing values by column
missing_values_by_column <- colSums(is.na(HeartRiskData))
cat("Missing values by column:\n")
print(missing_values_by_column)

# Install and load VIM package for visualization
library(VIM)

# Visualize missing data pattern
aggr(HeartRiskData, col = c("navy", "yellow"), numbers = TRUE, sortVars = TRUE, 
     labels = names(HeartRiskData), cex.axis = 0.7, gap = 3, ylab = c("Missing data", "Frequency"))
```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load necessary library
library(caret)

# Set the seed for reproducibility
set.seed(123)

# Split the dataset into training (80%) and testing (20%) sets
train_index <- createDataPartition(HeartRiskData$Risk_Level, p = 0.8, list = FALSE)

# Create training and testing datasets
train_data <- HeartRiskData[train_index, ]
test_data <- HeartRiskData[-train_index, ]

# Check the dimensions of the training and testing datasets
cat("Training data dimensions:", dim(train_data), "\n")
cat("Testing data dimensions:", dim(test_data), "\n")
```

## Cross-validation
```{r Cross-validation}
# Load the caret package
library(caret)

# Set up 10-fold Cross-Validation
ctrl_10fold <- trainControl(method = "cv", number = 10)


# Train a model using Random Forest with 10-fold CV
rf_model_cv <- train(Risk_Level ~ Age + Gender + Heart_rate + Systolic_blood_pressure + Diastolic_blood_pressure + 
                       Blood_sugar + CK_MB + Troponin + Result, 
                     data = HeartRiskData, 
                     method = "rf", 
                     trControl = ctrl_10fold)

# Print model results
print(rf_model_cv)
```

## Training Different Models
```{r Different Models}
# Load necessary libraries
library(caret)
library(e1071)  # for SVM if not already installed

# Set up 10-fold Cross-Validation
ctrl <- trainControl(method = "cv", number = 10)

# Random Forest Model
set.seed(123)
rf_model <- train(Risk_Level ~ Age + Gender + Heart_rate + Systolic_blood_pressure + 
                    Diastolic_blood_pressure + Blood_sugar + CK_MB + Troponin + Result,
                  data = HeartRiskData,
                  method = "rf",
                  trControl = ctrl)

# Print Random Forest Results
print(rf_model)


# k-Nearest Neighbors Model
set.seed(123)
knn_model <- train(Risk_Level ~ Age + Gender + Heart_rate + Systolic_blood_pressure + 
                     Diastolic_blood_pressure + Blood_sugar + CK_MB + Troponin + Result,
                   data = HeartRiskData,
                   method = "knn",
                   trControl = ctrl)

# Print KNN Results
print(knn_model)


# Support Vector Machine Model (Radial Basis Function Kernel)
set.seed(123)
svm_model <- train(Risk_Level ~ Age + Gender + Heart_rate + Systolic_blood_pressure + 
                     Diastolic_blood_pressure + Blood_sugar + CK_MB + Troponin + Result,
                   data = HeartRiskData,
                   method = "svmRadial",
                   trControl = ctrl)

# Print SVM Results
print(svm_model)
```

## Performance Comparison
```{r Performance Comparison}
# Load caret if not already loaded
library(caret)

# Compare the three models using resamples
model_comparison <- resamples(list(
  RandomForest = rf_model,
  KNN = knn_model,
  SVM = svm_model
))

# Summary of the model performance metrics
summary(model_comparison)

# Visual comparison plots
bwplot(model_comparison)   # Boxplot of accuracy, kappa, etc.
dotplot(model_comparison)  # Dotplot comparison
```

## Saving Model
```{r Saving Model}
# Create a folder if it doesn't exist
if (!dir.exists("./models")) {
  dir.create("./models")
}

# Save the best model (Random Forest)
saveRDS(rf_model, "./models/best_rf_model.rds")

# To load the saved model later
loaded_best_rf_model <- readRDS("./models/best_rf_model.rds")

# Example of making a prediction with the loaded model
new_data <- data.frame(
  Age = 55,
  Gender = factor(1, levels = levels(HeartRiskData$Gender)),
  Heart_rate = 72,
  Systolic_blood_pressure = 140,
  Diastolic_blood_pressure = 85,
  Blood_sugar = 180,
  CK_MB = 2.5,
  Troponin = 0.05,
  Result = factor("negative", levels = levels(HeartRiskData$Result))
)

# Predict the Risk_Level
predicted_risk <- predict(loaded_best_rf_model, newdata = new_data)

# Print the prediction
print(predicted_risk)

```

